Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-15 writeup/conStruct.tex
gid edits

%% added PLR
\newif\ifsubmissionversion
\submissionversiontrue
\newif\ifincludesupplement
\includesupplementtrue
\newif\iffloatsatend
\floatsatendfalse

@@

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

@@

\linenumbers

\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, reasonable people can disagree about whether two 
populations are separate species, because the 
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
The issue of identifying meaningful biological subunits 
extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long-distance migration. 
Both discrete and continuous components
are required to accurately describe most species' patterns of genetic relatedness.

@@

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \cite{CavalliSforza1975, treemix},
dimensionality-reduction approaches \cite{meirmans2009genodive}, 
such as principal components analysis 
\cite{menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
(e.g., \cite{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}), among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\cite{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster memberships \cite{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

\emph{Isolation by distance} refers to a pattern of increasing genetic differentiation
with geographic separation,
which occurs when geographically restricted dispersal allows
genetic drift to build up differentiation between distant locations
\cite{Wright1943}. 
Theoretical work,
mostly derived from ``stepping-stone'' models 
\cite{kimura1964stepping,sawyer1976stepping,shiga1984stepping},
gives us some analytical predictions for isolation by distance
\cite{malecot1969mathematics,Slatkin1985,epperson2003geographical}, 
but substantial work remains to be done \cite{barton2002neutral,barton2013modelling}.
% IBD has been modeled in both spatially continuous frameworks 
% in which an individual's parents are found within some radius of that individual's location,
% and discrete, ``stepping-stone'' frameworks \cite{kimura1964stepping},
% in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that IBD is very widespread in nature \cite{meirmans2012,Sexton_etal_2014}.

@@

To address this, we set out to develop
a model-based clustering method that, when possible, uses isolation by distance 
to explain observed genetic variation.
With an explicit spatial component, discrete population structure need only be invoked when genetic differentiation 
in the data deviates significantly from that expected given geographic separation.
In this paper, 
% we present a statistical method for simultaneously 
% modeling continuous and discrete patterns of population structure.
we model genetic variation in genotyped individuals as 
partitioned within or admixed across a specified number of discrete layers,
within each of which relatedness decays as a parametric function of the distance between samples.
We also implement a cross-validation approach for comparing and selecting models across different numbers of layers/clusters,
and we demonstrate the utility of our approach using both simulated and empirical data.
The implementation of this method, \texttt{conStruct} (for ``\emph{con}tinuous \emph{struct}ure"), 
is available as an R package at 
\href{https://github.com/gbradburd/conStruct}{https://github.com/gbradburd/conStruct}.

@@

This definition of covariance differs from the usual ``genetic covariance'' \cite{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eq.\ \eqref{allelic_covariance} further in \secref{allelic_cov}.
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $\pi_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 \pi_{i,j})/4$.
Therefore, 
this allelic covariance is more affected in shape by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if they are likely to contain a large percentage of errors.

@@

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
and the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the layer.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \cite{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
% (Analytical theory \cite{barton2002neutral} predicts the decay should be proportional to $\exp{-D}/\sqrt{D}$.)
The $\phi^{(k)}$ is a parameter that describes the background covariance within the layer. 
If two samples draw 100\% of their ancestry from layer $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the layer.
The ``shared drift'' parameter $\phi^{(k)}$ is analogous to 
the branch length connecting the $k$th population to the population ancestral to all modeled
layers (see for example \cite{patterson_ancient_2012,peter_fstats}),
although they cannot be directly compared because 
we are modeling the allelic, rather than genetic, covariance. 
In \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

@@

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;|\; \widehat{\Omega} \right) 
    \propto
    P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
    P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors 
except for $\alpha_2$, which is uniform on $(0,2)$, 
and $w$, for which we use an independent Dirichlet of dimension $K$ for each individual
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \cite{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \cite{R} called \texttt{conStruct} 
(for ``continuous structure") that functions as a wrapper around this inference machinery.

@@

We use cross-validation (similar in spirit to \cite{ADMIXTURE_xval}) to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

Cross-validation provides a valuable summary of how much explanatory power
is added by spatial structure within each layer, and each additional layer.
However, we remind users that ``statistical significance does not imply real-world significance'',
and so small but statistically significant differences between models
should probably not be relied on too strongly.

@@

With these simulations,
spatial \texttt{conStruct} does not create spurious discrete groupings when there are none:
Figures \ref{simK1_pies}, \ref{simK2_sp_pies}, and \ref{simK3_sp_pies} 
show that subsequent layers beyond the number used for simulation are unused.
When data are simulated with $K=1$ but analyzed with $K>1$ 
the layers are present, but contribute very little to any population.
Even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
Moreover, the method infers the true admixture proportions with high accuracy, tight precision, and good coverage 
(Figs \ref{simK2_adprop_fit} and \ref{simK3_adprop_fit}).

In contrast, the nonspatial model describes geographic variation
using gradients of admixture between more and more discrete clusters
to better approximate the continuous, spatial patterns of relatedness
(depicted in Figs \ref{simK1_pies} and \ref{simK1_nsp_pies}).
% as $K$ in the model increases, 
% the sampled area is increasingly divvied up between clusters 
% with ``centers of mass" in the $K$ geographical extremes of the simulated space.
% At $K=2$, the clusters are centered in opposite corners on the map; 
% at $K=3$, in the lower two corners and the top half; 
% at $K=4$, all four corners, and so on.
The fastSTRUCTURE results are qualitatively similar, as shown in Figs \ref{fastStr_simK1}-\ref{fastStr_simK3}.
Each nonspatial cluster is in truth genetically more similar within itself
than it is to other clusters,
but we know that these boundaries are arbitrary,
because the data were simulated with no such boundaries.

@@

To demonstrate the utility of this method, we analyzed empirical population genomic data from two systems:
a contact zone between two poplar species in northwestern North America,
and a large North American sample of black bears.

@@

Across multiple replicate runs, 
both nonspatial \texttt{conStruct} and fastSTRUCTURE displayed 
the successive partitioning of space and the clines of admixture seen in the simulation results.
The details of each were somewhat different 
(e.g. Fig \ref{populus_nsp_pies} vs. Fig \ref{populus_fastStr}), 
and also differed across the replicate analyses.

@@

The results partition the sampled bears into two main groups:
one (red) to the east of the Rocky Mountains, 
which also occurs in Alaska,
the other primarily west of the Rockies (blue).
The disjointed range of the blue layer likely reflects the fact
that Canada was not sampled, 
and so the red layer may extend
through the intervening (unsampled) northern Great Plains and Canadian Shield, 
with the blue layer presumably then stretching up into British Columbia.

@@

\section*{Allelic covariance}\label{allelic_cov}
To see how to arrive at our empirical variance-covariance expression, Eq.\ \eqref{allelic_covariance},
pick a random locus and
let $A$ and $B$ be randomly drawn alleles at that locus from populations $i$ and $j$ respectively.
If $i=j$, then these are chosen \emph{without} replacement.
Suppose these are each coded as `0' or `1` (where `0' denotes a reference allele),
but we randomly ``flip'' this coding, so that we let $X=A$ and $Y=B$ with probability 1/2,
but otherwise we let $X=1-A$ and $Y=1-B$.
We then let $\widehat{\Omega}_{i,j} = \cov[X,Y]$ ; 
sampling without replacement makes the statistic insensitive to sample size,
and random allele flipping makes it independent of the choice of reference allele.
By conditioning on the flip,
and using the fact that $\E[X] = \E[Y] = 1/2$,
Eq.\ \eqref{allelic_covariance} comes from the calculation that
\begin{align}
\cov[X,Y] &= \E[XY] - 1/4 = (1/2)\E[AB + (1-A)(1-B)] - 1/4\\\notag
&= \E[(A-1/2)(B-1/2)]
\end{align}
The factor of $\delta_{i,j}/(N_i-1)$ comes from the fact that if $i=j$,
then $\E[AB]$ is the probability that both alleles drawn without replacement are reference,
which is $f_i^2 n/(n-1)$.

@@

We employ Monte Carlo cross-validation approach for model comparison \cite{picard1984}. 
This procedure generates a mean predictive accuracy for each model and each value of $K$, 
as well as a confidence interval around that mean,
which can then be used for model comparison or selection.
Briefly, we follow the following procedure:
\begin{enumerate}
\item For each of $X$ replicates:
	\begin{enumerate}
		\item partition the allele frequency data into a 90\% ``training'' partition ($F^x_1$) and a 10 \%``testing" partition ($F^x_2$) \label{partition}
		\item run our inference procedure on the training partition to estimate model parameters $\theta_{mk}$ for: \label{inference}
			\begin{enumerate}
				\item $m$: the spatial and the nonspatial model
				\item $k$: the number of layers/clusters 1 through $K$
			\end{enumerate}	
	\item calculate the mean log likelihood of the testing data partition, 
	over the posterior distribution of training-estimated parameters for each model 
	($\bar{\mathcal{L}}(F^x_2 \mid \theta_{mk})$, henceforth $\bar{\mathcal{L}}_{xmk}$) \label{lnL}
	\item generate standardized mean log likelihoods, $\mathcal{Z}_{xmk}$, across models: \label{standardize}
		\begin{enumerate}
			\item identify the highest mean log likelihood, $\bar{\mathcal{L}}^\text{max}_{xmk}$
			\item subtract $\bar{\mathcal{L}}^\text{max}_{xmk}$ from $\bar{\mathcal{L}}_{xmk}$ across all models,
				such that the standardized log likelihood, $\mathcal{Z}_{xmk}$, of the best model is 0,
				and less than 0 for all inferior models. 
		\end{enumerate}
	\end{enumerate}
\item For each model (i.e., each combination of $m$ and $k$) calculate 
	the mean ($	\bar{\mathcal{Z}}_{mk} $) standardized log likelihood of the testing data partition across $X$ replicates, 
	as well its standard error ($SE_{\bar{\mathcal{Z}}_{mk}}$) and
        95\% confidence interval ($\bar{\mathcal{Z}}_{mk} \pm 1.96 \times SE_{\bar{\mathcal{Z}}_{mk}}$).
\end{enumerate}

@@

\newpage
\begin{figure}
	\centering
		{\includegraphics[width=\textwidth]{figs/sims/simK1_nsp_pies.pdf}}
	\caption{
	Map of admixture proportions estimated using a nonspatial model for $K=2$ through 7.
	The data were simulated using one layer with nearest-neighbor symmetric migration between demes.
    }\label{simK1_nsp_pies}
\end{figure}

>>>>

%% added PLR
\newif\ifsubmissionversion
%\submissionversionfalse
\submissionversiontrue
\newif\ifincludesupplement
\includesupplementtrue
\newif\iffloatsatend
\floatsatendfalse

@@

% Text layout
\ifsubmissionversion
    \raggedright
    \setlength{\parindent}{0.5cm}
    \textwidth 5.25in 
    \textheight 8.75in
\else
    \usepackage{fullpage}
\fi

@@

\ifsubmissionversion
\linenumbers
\fi

\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, reasonable people can disagree about whether two 
populations are separate species because the 
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
The issue of identifying meaningful biological subunits 
extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long-distance migration. 
Both discrete and continuous components
are required to accurately describe most species' patterns of genetic relatedness.

@@

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \cite{CavalliSforza1975, treemix},
dimensionality-reduction approaches, 
such as principal components analysis 
\cite{meirmans2009genodive,menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
(e.g., \cite{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}), among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\cite{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster memberships \cite{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

\emph{Isolation by distance} (IBD) refers to a pattern of increasing genetic differentiation
with geographic separation,
which occurs when geographically restricted dispersal allows
genetic drift to build up differentiation between distant locations
\cite{Wright1943}. 
Theoretical work,
mostly derived from ``stepping-stone'' models 
\cite{kimura1964stepping,sawyer1976stepping,shiga1984stepping},
gives us some analytical predictions for isolation by distance
\cite{malecot1969mathematics,Slatkin1985,epperson2003geographical}, 
but substantial work remains to be done \cite{barton2002neutral,barton2013modelling}.
% IBD has been modeled in both spatially continuous frameworks 
% in which an individual's parents are found within some radius of that individual's location,
% and discrete, ``stepping-stone'' frameworks \cite{kimura1964stepping},
% in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that IBD is very widespread in nature \cite{meirmans2012,Sexton_etal_2014}.

@@

To address this, we set out to develop
a model-based clustering method that, when possible, uses isolation by distance 
to explain observed genetic variation.
With an explicit spatial component, discrete population structure need only be invoked when genetic differentiation 
in the data deviates significantly from that expected given geographic separation.
In this paper, 
% we present a statistical method for simultaneously 
% modeling continuous and discrete patterns of population structure.
we model genetic variation in genotyped individuals as 
partitioned within or admixed across a specified number of discrete layers,
within each of which relatedness decays as a parametric function of the distance between samples.
We also implement a cross-validation approach for comparing and selecting models across different numbers of layers,
and we demonstrate the utility of our approach using both simulated and empirical data.
The implementation of this method, \texttt{conStruct} (for ``\emph{con}tinuous \emph{struct}ure"), 
is available as an R package at 
\href{https://github.com/gbradburd/conStruct}{https://github.com/gbradburd/conStruct}.

@@

This definition of covariance differs from the usual ``genetic covariance'' \cite{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eq.\ \eqref{allelic_covariance} further in \secref{allelic_cov}.
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $\pi_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 \pi_{i,j})/4$.
Therefore, 
this allelic covariance is more affected in shape by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if they are likely to contain a large percentage of errors \cite{linck_battey2017}.

@@

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
and the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the layer.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \cite{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
% (Analytical theory \cite{barton2002neutral} predicts the decay should be proportional to $\exp{-D}/\sqrt{D}$.)
The $\phi^{(k)}$ is a parameter that describes the background covariance within the layer. 
If two samples draw 100\% of their ancestry from layer $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to the background level $\phi^{(k)}$ within the layer.
The ``shared drift'' parameter $\phi^{(k)}$ is analogous to 
the branch length connecting the $k$th population to the population ancestral to all modeled
layers (see for example \cite{patterson_ancient_2012,peter_fstats}),
although they cannot be directly compared because 
we are modeling the allelic, rather than genetic, covariance. 
In \secref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

@@

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;|\; \widehat{\Omega} \right) 
    \propto
    P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
    P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors 
except for $\alpha_2$, which is uniform on $(0,2)$, 
and $w$, for which we use an independent Dirichlet of dimension $K$ for each sample
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \cite{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \cite{R} called \texttt{conStruct} 
(for ``\emph{con}tinuous \emph{struct}ure") that functions as a wrapper around this inference machinery.

@@

We use cross-validation (similar in spirit to \cite{ADMIXTURE_xval}) to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} for more on our cross-validation procedure).
In general, larger values of $K$ allow the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
as above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

Cross-validation provides a valuable summary of how much explanatory power
is added by spatial structure within each layer, and each additional layer.
However, we remind users that ``statistical significance does not imply real-world significance','
and so small but statistically significant differences between models
should probably not be relied on too strongly.

@@

With these simulations,
spatial \texttt{conStruct} does not create spurious discrete groupings when there are none:
Figures \ref{simK1_pies}, \ref{simK2_sp_pies}, and \ref{simK3_sp_pies} 
show that subsequent layers beyond the number used for simulation are unused.
When data are simulated with $K=1$ but analyzed with $K>1$, 
the layers are present, but contribute very little to any population.
Even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
Moreover, the method infers the true admixture proportions with high accuracy, tight precision, and good coverage 
(Figs \ref{simK2_adprop_fit} and \ref{simK3_adprop_fit}).

In contrast, the nonspatial model describes geographic variation
using gradients of admixture between more and more discrete clusters
to better approximate the continuous, spatial patterns of relatedness
(depicted in Figs \ref{simK1_pies} and \ref{simK1_nsp_pies}).
% as $K$ in the model increases, 
% the sampled area is increasingly divvied up between clusters 
% with ``centers of mass" in the $K$ geographical extremes of the simulated space.
% At $K=2$, the clusters are centered in opposite corners on the map; 
% at $K=3$, in the lower two corners and the top half; 
% at $K=4$, all four corners, and so on.
The fastSTRUCTURE results are qualitatively similar, as shown in Figs \ref{fastStr_simK1}-\ref{fastStr_simK3}.
Each nonspatial cluster is in truth genetically more similar within itself
than it is to other clusters,
but we know that these boundaries are arbitrary,
because the data were simulated without them.

@@

To further demonstrate the utility of this method, 
we also analyzed empirical population genomic data from two systems:
a contact zone between two poplar species in northwestern North America,
and a large North American sample of black bears.

@@

Across multiple replicate runs, 
both nonspatial \texttt{conStruct} and fastSTRUCTURE displayed 
the successive partitioning of space and the clines of admixture seen in the simulation results.
The details of each were somewhat different 
(Fig \ref{populus_nsp_pies} vs. Fig \ref{populus_fastStr}), 
and also differed across the replicate analyses.

@@

The results partition the sampled bears into two main groups:
one (red) to the east of the Rocky Mountains, 
which also occurs in Alaska,
the other primarily west of the Rockies (blue)  (Fig \ref{bear_K3}a).
The disjointed range of the blue layer likely reflects the fact
that Canada was not sampled, 
and so the red layer may extend
through the intervening (unsampled) northern Great Plains and Canadian Shield, 
with the blue layer presumably then stretching up into British Columbia.

@@

\section*{Allelic covariance}\label{allelic_cov}
To see how to arrive at our empirical variance-covariance expression (Eq.\ \eqref{allelic_covariance})
pick a random locus and
let $A$ and $B$ be randomly drawn alleles at that locus from populations $i$ and $j$ respectively.
If $i=j$, then these are chosen \emph{without} replacement.
Suppose these are each coded as `0' or `1` (where `0' denotes a reference allele),
but we randomly ``flip'' this coding, so that we let $X=A$ and $Y=B$ with probability 1/2,
but otherwise we let $X=1-A$ and $Y=1-B$.
We then let $\widehat{\Omega}_{i,j} = \cov[X,Y]$ ; 
sampling without replacement makes the statistic insensitive to sample size,
and random allele flipping makes it independent of the choice of reference allele.
By conditioning on the flip,
and using the fact that $\E[X] = \E[Y] = 1/2$,
Eq.\ \eqref{allelic_covariance} comes from the calculation that
\begin{align}
\cov[X,Y] &= \E[XY] - 1/4 = (1/2)\E[AB + (1-A)(1-B)] - 1/4\\\notag
&= \E[(A-1/2)(B-1/2)]
\end{align}
The factor of $\delta_{i,j}/(N_i-1)$ comes from the fact that if $i=j$,
then $\E[AB]$ is the probability that both alleles drawn without replacement are reference,
which is $f_i^2 n/(n-1)$.

@@

We employ Monte Carlo cross-validation approach for model comparison \cite{picard1984}. 
This procedure generates a mean predictive accuracy for each model and each value of $K$, 
as well as a confidence interval around that mean,
which can then be used for model comparison or selection.
Briefly, we follow the following procedure:
\begin{enumerate}
\item For each of $X$ replicates:
	\begin{enumerate}
		\item partition the allele frequency data into a 90\% ``training'' partition ($F^x_1$) and a 10 \%``testing" partition ($F^x_2$) \label{partition}
		\item run our inference procedure on the training partition to estimate model parameters $\theta_{mk}$ for: \label{inference}
			\begin{enumerate}
				\item $m$: the spatial and the nonspatial model
				\item $k$: the number of layers/clusters 1 through $K$
			\end{enumerate}	
	\item calculate the mean log likelihood of the testing data partition  
	over the posterior distribution of training-estimated parameters for each model 
	($\bar{\mathcal{L}}(F^x_2 \mid \theta_{mk})$, henceforth $\bar{\mathcal{L}}_{xmk}$) \label{lnL}
	\item generate standardized mean log likelihoods, $\mathcal{Z}_{xmk}$, across models: \label{standardize}
		\begin{enumerate}
			\item identify the highest mean log likelihood, $\bar{\mathcal{L}}^\text{max}_{xmk}$
			\item subtract $\bar{\mathcal{L}}^\text{max}_{xmk}$ from $\bar{\mathcal{L}}_{xmk}$ across all models,
				such that the standardized log likelihood, $\mathcal{Z}_{xmk}$, of the best model is 0,
				and less than 0 for all inferior models. 
		\end{enumerate}
	\end{enumerate}
\item For each model (i.e., each combination of $m$ and $k$) calculate 
	the mean ($	\bar{\mathcal{Z}}_{mk} $) standardized log likelihood of the testing data partition across $X$ replicates, 
	as well its standard error ($SE_{\bar{\mathcal{Z}}_{mk}}$) and
        95\% confidence interval ($\bar{\mathcal{Z}}_{mk} \pm 1.96 \times SE_{\bar{\mathcal{Z}}_{mk}}$).
\end{enumerate}

@@

\begin{figure}[h]
	\centering
		{\includegraphics[width=\textwidth]{figs/sims/simK1_nsp_pies.pdf}}
	\caption{
	Map of admixture proportions estimated using a nonspatial model for $K=2$ through 7.
	The data were simulated using one layer with nearest-neighbor symmetric migration between demes.
    }\label{simK1_nsp_pies}
\end{figure}
